Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-06-01 manuscript/main.qmd
final manuscript corrections

Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018]. Recent advances in the synthesis of palaeoclimate data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, yet a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

@@

We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. We choose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018]. Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$. For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples from the continental shelf. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained. For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 &deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

@@

This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$--> but is drawn towards the overall logistic regression curve, i.e. towards $\nu_j$. The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e. $\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low. In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.

@@

**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to limit very abrupt SST changes across latitudes on a global scale. A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$. We informed the prior distributions on M and B based on a provisional model run with the modern SST data.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992]. We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples. To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1&deg; latitude and calculated the $R^2$ based on these values. We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values. Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.

@@

The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude). To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of at the poles. Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [@Evans2018].

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, and extreme greenhouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 2.5% and 97.5% credible intervals from the 1,000 model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$). The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only very slightly higher at `r modern_gradient$"50%"`&deg;C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).

@@

Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model. The ribbons (purple and turquoise shading) depict the 95% credible intervals of these gradients, respectively. Points within the plot depict the geochemical (e.g. &delta;^18^O) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted at the mean temperature values of their respective normal distributions.
:::

@@

Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data, even with small sample sizes (*n* = 10 - 20) and patchy sampling distributions (Fig. 2). This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g. @Bijl2009; @Tierney2017], which can fit only specific types of gradients. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020]. This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns without the option of including additional knowledge (e.g. that latitudinal temperature gradients should be broadly negative).

@@

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012]. One major limitation to these methods is that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@kleypas1999]. Although coral reefs are threatened by warming sea surface temperatures today [@Hoegh2011], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, with few early Eocene coral reefs occurring in low latitudes [@Zamagni2012]. The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irradiance, rather than temperature [@Muir2015]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for coral reefs, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of 13&deg;C [@Inglis2020]. Our polar circle to equatorial gradient estimate is lower than most previous estimates at 7.8&deg;C, although the 95% credible interval extends up to 13.7&deg;C and thus overlaps earlier estimates based on shallow water proxies. The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.

@@

Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 -- 37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including $TEX_{86}$ [@Inglis2020]. Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5&deg;C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`&deg;C.

>>>>

Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018; @Tierney2020]. Recent advances in the synthesis of palaeoclimate data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, yet a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

@@

We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. We choose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018]. Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1\; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples from the continental shelf. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained. For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 &deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene -- early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

@@

This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$--> but is drawn towards the overall logistic regression curve, i.e. towards $\nu_j$. The pull towards $\nu_j$ tends to be strong when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e. $\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low. In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent reconstructed temperatures.

@@

**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to limit very abrupt SST changes across latitudes on a global scale. A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$. We informed the prior distributions on $M$ and $B$ based on a provisional model run with the modern SST data.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992]. We then randomly sampled (1,000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples. To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1&deg; latitude and calculated the $R^2$ based on these values. We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values. Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.

@@

The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude). To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of the temperature at the poles. Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as has been done in some earlier studies [@Evans2018].

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, and extreme greenhouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$). The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).

@@

Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model. The purple ribbon (shading) depict the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible. Points within the plot depict the geochemical (e.g. TEX~86~) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted at the mean temperature values of their respective normal distributions.
:::

@@

Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data, even with small sample sizes (*n* = 10 -- 20) and patchy sampling distributions (Fig. 2). This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g. @Bijl2009; @Tierney2017], which can fit only specific types of gradients. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020]. This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns, without the option of including additional knowledge (e.g. that latitudinal temperature gradients should be broadly negative).

@@

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012]. One major limitation to these methods is that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@kleypas1999]. Although coral reefs are threatened by warming sea surface temperatures today [@Hoegh2011], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, with few early Eocene coral reefs occurring in low latitudes [@Zamagni2012]. The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irradiance, rather than temperature [@Muir2015]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig.&nbsp;S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for coral reefs, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our polar circle to equatorial gradient estimate is lower than most previous estimates at 7.8&deg;C, although the 95% credible interval extends up to 13.7&deg;C and thus overlaps earlier estimates based on shallow water proxies. The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.

@@

Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 -- 37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including TEX~86~ [@Inglis2020]. Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5&deg;C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`&deg;C.
